Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-02-05 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Added line breaks. No content change.
more edits...
Section 5.2 and Appendix E: minor changes.

Collaborative funding can be important for RSE groups, to ensure that research
leadership remains with the domain scientists. At NCAR, university partnerships
are required for submission of proposals, so collaboration is an essential part
of grant submission, and this will tend to bring developers and scientists
together. The UCL group also follows this approach, with all bids requiring an
academic collaborator.

@@

One model that has been helpful in finding funding for RSE groups is the use of
funds left over on research grants when RAs have left prematurely -- PIs like
this arrangement as it is hard to find good staff for short-term positions, so
having a pool of research programming staff on hand resolves this problem. We
recommend that funders give explicit guidance to grant holders and institutions
that such arrangement is favorable. Framework agreements permitting this to go
ahead without checking back every time with funders and/or grant panels would
further smooth this. (This also provides more stable jobs for those who hold
these skills, but arguments about making life nicer for postdocs will not help
persuade funders or PIs!)

There is some question about the most effective duration and percent of full
time for a programmer's work on a project. At least three months is necessary
for the programmer to read into the science (RSEs must not become so disengaged
from research that they do not have time to read a few papers -- this will result
in code which does not meet scientific needs), but too long could result in an
RSE losing their flexibility, becoming so engaged in one project that when that
project ends, they find it hard to transfer. For this reason, we also recommend
that 40 percent is ideal; two projects per developer, with some time for
training and infrastructure work. Having two developers per project seems to be ideal,
in the sense that software development is enhanced by two pairs of eyes.

@@

In the UK, another source of funding for research software has been the
Collaborative Computational Projects (CCPs): domain specific communities put
forward proposals that are a priority of the community as a whole, for example,
biosimulation or plasma physics. These bodies act as custodians of community
codes, and a central team also provides software engineering support.

@@

It can be a more difficult to help leading domain scientists see the value of
engineering effort than those in their teams who are forced to work with
difficult-to-use or unreliable software tools, as they do not see the pain.
Perhaps a version of ``software carpentry'' targeted at those PIs who are
awarded or apply for software-intensive grants could be valuable here.

@@

Financial support for an initial conference to bring together research software
groups to form an organization and create a resource sharing structure would
help to further this campaign. Funding to conduct and analyze a survey could
also be quite useful as knowing where we stand today, and what models are in use
could fuel the ideas for further development of developers in this category.

>>>>

Collaborative funding can be crucial to RSE groups, to ensure that research
leadership remains with the domain scientists. As an example, at NCAR, university partnerships
are required for submission of proposals, so collaboration is an essential part
of grant submission, and this will tend to bring developers and scientists
together. The UCL group also follows this approach, with all bids requiring an
academic collaborator.

@@

One model that has been helpful in finding funding for RSE groups is the use of
funds left over on research grants when RAs have left prematurely -- PIs like
this arrangement as it is hard to find good staff for short-term positions, so
having a pool of research programming staff on hand resolves this problem. We
recommend that funders give explicit guidance to grant holders and institutions
that such an arrangement is favorable. Framework agreements permitting this to go
ahead without checking back every time with funders and/or grant panels would
further smooth this. (This also provides more stable jobs for those who hold
these skills, but arguments about making life nicer for postdocs will not help
persuade funders or PIs!)

There is some question about the most effective duration and percentage of full
time for a programmer's work on a project. At least three months is necessary
for the programmer to read into the science (RSEs must not become so disengaged
from research that they do not have time to read a few papers -- this will result
in code which does not meet scientific needs), but too long could result in an
RSE losing their flexibility, becoming so engaged in one project that when that
project ends, they find it hard to transfer. For this reason, we also recommend
that 40 percent is ideal; two projects per developer, with some time for
training and infrastructure work. Having two developers per project seems to be ideal,
in the sense that software development is enhanced by two pairs of eyes.

@@

In the UK, another source of funding for research software is the
Collaborative Computational Projects (CCPs): domain specific communities put
forward proposals that are a priority of the community as a whole, for example,
biosimulation or plasma physics. These bodies act as custodians of community
codes, and a central team also provides software engineering support.

@@

It can be more difficult to help leading domain scientists see the value of
engineering effort than those in their teams who are forced to work with
difficult-to-use or unreliable software tools, as they do not see the pain.
Perhaps a version of ``software carpentry'' targeted at those PIs who are
awarded or apply for software-intensive grants could be valuable here.

@@

Financial support for an initial conference that brings together research software
groups to form an organization and create a resource sharing structure would
help to further this campaign. Funding to conduct and analyze a survey could
also be quite useful as knowing where we stand today, and what models are in use
could fuel the ideas for further development of developers in this category.

### 2016-02-05 WSSSPE3/report/wg_main_funding_spec_expert.tex
more review of changes
Section 5.2 and Appendix E: minor changes.

While development of new mechanisms in the current academic reward system is a
worthy aspirational goal, such a dramatic change in this structure does not seem
likely in a time scale relevant to this working group. Accordingly, our working
party sought alternative solutions that may be achievable within the context of
existing academic structures. The group felt that developing dedicated research
software engineering roles within the university, and finding stable funding for
those individuals is the most promising mechanism for creating a stable software
development staff.

@@

One solution is to fund dedicated software engineering roles for major research
software projects at national laboratories or other non-educational
institutions. This solution is in place and working well for many well used
scientific codebases. However, this strategy has limited application, as much of
the body of software is created and maintained in research universities.
Therefore, we argue that research institutions should develop hybrid
academic-technical tracks for this capacity, where employees in this track work
with more than one PI, rather than the traditional RA role within a single
group. This could be coordinated centrally, as a core facility, perhaps within
research computing organizations which have traditionally supported university
cyberinfrastructure, library organizations, or research offices. Alternatively,
these groups could be organizationally closer to research groups, sitting within
academic departments. The most effective model will vary from institution to
institution, but the mandate and ways of working should be similar.

>>>>

While developing new mechanisms in the current academic reward system is a
worthy aspirational goal, such a dramatic change in this structure does not seem
likely in a time scale relevant to this working group. Accordingly, our working
party sought alternative solutions that may be achievable within the context of
existing academic structures. The group felt that developing dedicated research
software engineering roles within the university, and finding stable funding for
those individuals is the most promising mechanism for creating a stable software
development staff.

@@

One solution is to fund dedicated software engineering roles for major research
software projects at national laboratories or other non-educational
institutions. This solution is in place and working well for many well-used
scientific codebases. However, this strategy has limited application, as much of
the body of software is created and maintained in research universities.
Therefore, we argue that research institutions should develop hybrid
academic-technical tracks for this capacity, where employees in this track work
with more than one PI, rather than the traditional RA role within a single
group. This could be coordinated centrally, as a core facility, perhaps within
research computing organizations which have traditionally supported university
cyberinfrastructure, library organizations, or research offices. Alternatively,
these groups could be organizationally closer to research groups, sitting within
academic departments. The most effective model will vary from institution to
institution, but the mandate and ways of working should be similar.
